---
title: "Worldviews and filter bubbles generated from World Value Survey statements by interacting agents using LLMs"
date: today
author:
  - name: Jan Lorenz
    id: jl
    orcid: 0000-0002-5547-7848
    email: jlorenz@constructor.university
    affiliation: 
      - name: Constructor University
        city: Bremen
        state: Germany
        url: https://www.constructor.university
  - name: Erkan Gunes
    id: eg
    orcid: 0009-0005-8106-3314
    email: eguenes@constructor.university
    affiliation: 
      - name: Constructor University
        city: Bremen
        state: Germany
        url: https://www.constructor.university
abstract: > 
  We present the concept of reproducing an agent-based model of opinion dynamics by replacing an abstract numerical opinion space with large-language models (LLMs), capable of generating human-like statements in natural language, to decide how agents change their opinions. We aim to validate by computer simulation whether the mechanism of filter bubble creation through social media will also show up in this more realistic setting. We also discuss different ways to implement parts of the original model using LLM agents and the potential challenges we may face integrating LLMs into an agent-based model of opinion dynamics. 
keywords:
  - agent-based models
  - large-language models
  - filter bubbles
  - opinion dynamics
  - political space
  - polarization
# citation: 
#   container-title: Journal of Psychoceramics
#   volume: 1
#   issue: 1
#   doi: 10.5555/12345678
funding: 'The work benefited from funding in the Horizon Europe project "European City Squared: Computational Social Science Simulation for Democracy" (EuropeanCity2 Grant agreement ID: 101178170, doi: [10.3030/101178170](https://doi.org/10.3030/101178170)).'
---

::: {.text-center}
[**This is work in progress and not yet ready for citation.**]{.text-red}  
[Please contact the authors if you want to use or discuss this work.]{.text-red}
:::

```{r}
#| include: false
library(tidyverse)
library(arrow)
library(readxl)
library(glue)
library(patchwork)
library(ggh4x) # Hacks for ggplot, here to make facets different widths
library(gt)
library(ggiraph)

memory <- read_parquet("parquet/memory.parquet")
runs_core <- read_parquet("parquet/runs_core.parquet")
statements_meta <- read_parquet("parquet/statements_meta.parquet")
wvs <- read_parquet("parquet/wvs.parquet")
load("parquet/PCAall.Rdata")
PC1s <- read_parquet("PC1s.parquet")
runs_annotations <- read_xlsx("rawdata/runs_annotations.xlsx")
```


# Introduction

Opinion dynamics is one of the main topics in agent-based modeling of sociopolitical systems [@FlacheEtAl2017ModelsSocialInfluence; @LorenzEtAl2021IndividualAttitudeChange]. Models of opinion dynamics have been used to understand the mechanisms of societal radicalization and polarization, and the emergence of filter bubbles in various ways [@KeijzerMaes2022ComplexLinkFilter; @KeijzerEtAl2026ComputationalSocialScience]. 

A shortcoming of opinion dynamics models has always been that something as complex as the opinion of a human being has to be boiled down to one number (or a vector of a few numbers) in an abstract numerical opinion space such as the left-right political continuum. Of course, numerical opinion spaces also have applied use. Researchers measure the attitudes of individuals with specific or large-scale international (representative) surveys, where individuals self-assess their attitudes on (standardized) scales, e.g. with numbers from 0 to 10. In addition, actors in political discourse in the real world use spatial metaphors such as a politician "moving to the right, left, or center". So, political science has developed several methods for positioning political actors and texts in low-dimensional numerical opinion spaces. These methods range from expert assessments using theory-driven classification to data-driven approaches measuring positions in latent dimensions of higher-dimensional spaces using roll-call, co-voting data, survey data, or texts using dimensionality reduction and natural language processing. 

The well-known models of continuous opinion dynamics [see @Lorenz2025ExploringTheoryAgentbased] make use of such abstract one-(or low-)dimensional opinion spaces and define how individuals change their opinion on that scale when confronted with the opinions of others. However, an attitude change does not occur based on the exchange of numerical numbers but through verbalization of arguments [@BanischOlbrich2019OpinionPolarizationLearning], persuasive messages, and emotional emphasis, whether written or spoken language, or other media such as images. While it is plausible that the attitudes of individuals can be mapped reasonably well into a low-dimensional numerical space for descriptive purposes, it is less realistic that individual attitude change can be well operationalized solely based on just the exchange of numbers.

Large Language Models (LLMs) provide a new opportunity to build a new generation of opinion dynamics models. LLMs trained on large bodies of text seem to have embedded the meaning of the antagonisms and compatibilities of opinions as stated in written texts. 

In this paper, we build such a model inspired from the filter bubble model of  @GeschkeEtAl2019TriplefilterBubbleUsing. The main entities in the model are *agents* positioned in a two-dimensional numerical opinion space, and *statements* (called *bits of information*) positioned in the same opinion space. They are linked to a limited number of statements composing their *worldview*. They define their position in the opinion space by the average position of their worldview (in both dimension). Further on, agents maintain a social network of bidirectional friendship links.  Based on this *setup*, agents have a few *behavioral mechanisms* implemented. When they are presented with a new statement, either from a central media or their friends, they decide whether to *integrate* it into their worldview or not. If they integrate it they *drop* another statement. With a change of statements they *adjust* their position in opinion space to the new worldview average. They can *share* a statements of their worldview along all their friendship links, mimicking its posting on social media. 

The macroscopic *opinion landscape* of agents and statements in the model is thus a bipartite network of agents and statements, where agents are linked to the statements of their worldview, and statements are linked to the agents that have them in their worldview. In such an opinion landscape, a *bubble* is a set of agents whose worldviews have a high overlap of statements while the overlap with all other agents is low or non-existent. The model shows the following emergent mechanisms/phenomena:

- When new statement comes in from a central media source which every agent listens to, the opinion landscape develops a structure characterized by high and low density regions but the network remains fully connected without strong bubbles. 
- When new statement come additionally from friends, the opinion landscape develops a structure characterized by bubbles. How many bubbles emerge depends to a large extent in a parameter called *latitude of acceptance* which defines how far away a statement can be from the average position of an agent's worldview to be integrated. When this latitude is small, many small bubbles emerge. When it is large enough, only one large bubble emerges.

This model setup can be transformed into a *text-based setup* without a numerical opinion space. Statements transform from dots in two-dimensional numerical space to *textual statements*. Instead of linking to numerical positions, agents maintain a list of statements as their worldview. The behavioral mechanisms of integrating and dropping statements are performed using an *LLM prompt*. 

In this study we used a limited set of 100 statements extracted from the World Value Survey questionnaire. Each statement is directly related to one survey item and to a certain response (or response scale). This creates options for validation of emergening worldviews in LLM simulations with empirical survey responses.

The phenomena we want to study with this model are on different levels. 

On generative phenomena:

- Does the textual representation of opinions and the use of LLMs for integration decisions lead to the same phenomena of bubble formation as the numerical model? 
- Is the presence or absense of social posting of statements by friends similarily relevant for the emergence of bubbles?
- Do we see new phenomena emerging in the text-based model which were not present in the numerical model?

On interacting LLMs:

- Which statements do LLM agents prefer to integrate into their worldview?
- To which worldviews do LLM agents converge when they interact with each other?

On model validation:

- Do the worldviews emerging in the LLM-based model show a similar structure as the worldviews of individuals in the World Values Survey?

[*This list of research questions in not exhaustive and will be extended as we progress with the implementation and analysis of the model.*]{.red}




# Methods and Data

## Model

The implementation of the agent-based model using LLMs for agent's worldview integration decisions follows mostly our outline [@LorenzGunes2025FilterBubblesAgentbased]. 

*To do.*

Feed posting and social positing. 

In each time step all agents first received the same set of three randomly selected statements from the total set of statements and decide about integrating them. 
Then every agent selected a random statement from their worldview and shares it with all friends. When every agent shared. Each agent gathers the list of the shared statements of all friends and decides upon integration and dropping again. This is repeated for 100 time steps.  


*To Do*

This is an example prompt starting with an ever equal system prompt for all agents and then a user prompt with the same content but different statements for each agent:

> You are an individual with a worldview represented by a list of statements. You are presented with a list of new statements coming to your attention. Choose the single statement that best fits with your worldview, or reject all of them if no statement fits well with your worldview. Your list of statements is considered full when it contains 10 statements. Respond in JSON with two entries: 'selected_statement' must be the index number of the new statement you choose, or 'None' if you do not select any statement because they all do not fit well with your worldview. 'drop_statement' must be the index number of the statement you will drop from your list of worldview statements if your memory is full, or 'None' if you do not need to drop a statement. For example, if there are three new statements shown, then the value of the 'selected_statement' entry must be 1, 2, 3, or 'None'.
On a scale from 0 to 10, where 0 means you almost never accept new statements and 10 means you almost always accept new statements, your openmindedness is 3.
> 
> Here are the statements you can choose from:
>
> 1. "Government should take more responsibility to ensure everyone is provided for"
2. "Science and technology are making our lives healthier, easier, and more comfortable"
3. "We depend too much on science and not enough on faith"
>
> Here is the list of your existing worldview statements:
>
> 1. "Honest elections make an important difference in people’s lives"
2. "Homosexuality should be treated as totally normal"
3. "My most imporant personal goal is protecting freedom of speech"
4. "Political violence is never justifiable"
5. "The government should not monitor emails and Internet communication"
6. "In a democracy it is essential that women have the same rights as men"
7. "Freedom is more important than security"
8. "In a democracy it is essential that civil rights protect liberty from state oppression"
9. "Homosexual couples are as good parents as other couples."
10. "Abortion should be legal"


This particular prompt resulted in the outcome:

> Agent selected: "Science and technology are making our lives healthier, easier, and more comfortable"  
> Agent dropped: "The government should not monitor emails and Internet communication"


*To do.*

## Data

*To do.*

World Value Survey (WVS) data [@HaerpferEtAl2022WorldValuesSurvey]. 

Our selection of 100 statements from the WVS questionnaire is documented at [WVS questions for ABM-LLM](https://docs.google.com/spreadsheets/d/e/2PACX-1vTy7URowqvqpj1K4LfuwLgaNUTsxKOute6D6OSJbTY9C9DSz8eDbeib9mqwE83BltlVamscY3WsKK2V/pubhtml?gid=0&single=true). The WVS questions statements were selected to cover a wide range of topics and to be transformable into a worldview statement. For some questions more than one statement has been generated by the same question, for example taking the two opposing ends from an answering scale or different statements from a rank order question. In total the 100 statements were derived from `r (100-statements_meta$made_from_other_question |> sum(na.rm=T))` questions. So, still each statement is linked to a specific question in the WVS questionnaire, allowing for comparing the model's outcomes to the empirical survey data. 

*To do.*



# Results

The simulation runs we have generated can be explored in the [Network Explorer](networkexplorer.qmd). There agents and statements are visualized in a bipartite network using a layout computed dynamically with a version of the force-directed algorithm [@FruchtermanReingold1991GraphDrawingForcedirected]. Snapshots of simulation runs can be explored at time steps 5, 10, 20, 50, and 100. 

All runs were with 100 agents, the 100 statements from our World Value Survey pool and runing for 100 time steps. Simulations started with each agents having one random statement in memory. 

@fig-twobubbles, @fig-authbubble, and @fig-libbubble show the outcomes of typical simulation runs. In @fig-twobubbles a society with two *bubbles* of worldviews have emerged. Those bubbles are connected only by a few agent having statements from both sides in their memory and by only one statement being being equally often shared by agents of both bubbles (it is "I am very worried about a civil war" here). This is a good point to give a qualitative definition of a *bubble*: A group of agents which have very similar worldviews composed of a small set of statements, while these statements are also rarely shared by agents not in the bubble. 

@fig-libbubble and @fig-authbubble show simulation runs with the same parameters and the same initial conditions. The runs in the three Figures only differ in which random statments from the total pool of statements are in the feed postings and which statements are selected at random by agents for social posting. The actual LLM made decisions could also contribute further stochasticity to the outcome, as even identical prompts can deliver different decisions in differnt LLM calls. 

In @fig-libbubble and @fig-authbubble, the worldview of the same agent is highlighted. The societies in these examples are characterzied by essentially one big bubble. Therefore, the worldview shown is prototypical for the whole society. Interestingly, the worldviews are very different. @fig-libbubble shows a *liberal* bubble emphasizing freedom and equality in very abtract terms. The bubble in @fig-authbubble show more authoritarian statements, emphasizing security and defense against threats. Further analysis shows that also the two bubbles in @fig-twobubbles are the same liberal and authoritarian bubbles. Over all, these two bubbles dominate all simulation runs either as one or with both bubbles emerging.

*To Do: Explain what other statements are added in the memory 10 outcomes or when only feedposting is used (less bubbleness).*





::: {#fig-twobubbles fig-cap="Snapshot of a simulation run at $t=100$ with agents (blue) and the statements (red) they have in memory. Simulation with feed posting and social posting under openmindedness set to 1 and a maximal number of 5 statements in agents' memories. [Network Explorer: 260630-120838_A100_M5_tmax100_op1_FS_seed354_gpt](networkexplorer.html?net=260630-120838_A100_M5_tmax100_op1_FS_seed354_gpt_t100)" }
```{=html}
<iframe
  src="network-embed.html?file=networkjson/260630-120838_A100_M5_tmax100_op1_FS_seed354_gpt_t100.json&type=AS&zoom=1&spread=100&edgelen=10&attract=0.75"
  style="width:100%; height:300px; border:none;">
</iframe>
```
:::

![Snapshot of a simulation run at $t=100$ analog to @fig-twobubbles using the same initial condition but different random feed and social posting. Highlighted is the worldview of agent 30. [Network Explorer: 260630-101930_A100_M5_tmax100_op1_FS_seed354_gpt_t100](networkexplorer.html?net=260630-101930_A100_M5_tmax100_op1_FS_seed354_gpt_t100)](img/260630-101930_A100_M5_tmax100_op1_FS_seed354_gpt_t100.svg){#fig-libbubble}


![Snapshot of a simulation run at $t=100$ analog to @fig-twobubbles using the same initial condition but different random feed and social posting. Highlighted is the worldview of agent 30. [Network Explorer: 260630-141000_A100_M5_tmax100_op1_FS_seed354_gpt_t100](networkexplorer.html?net=260630-141000_A100_M5_tmax100_op1_FS_seed354_gpt_t100](img/260630-141000_A100_M5_tmax100_op1_FS_seed354_gpt_t100.svg){#fig-authbubble}





## Statement Ideology based on WVS

```{r}
load("parquet/PCAall.Rdata")
sttmts <- read_parquet("sttmts.parquet")
```

A principal component analysis of the 100 statements based on the responses of all respondents of the World Value Survey 2018 with complete data (N = `r nrow(PCAall$x)`) shows a clear left-right ideological dimension. The first principal component (PC1) explains 13.5% of the variance in the data and is highly correlated with the left-right self-placement of respondents (r = 0.82). The statements can be ordered along this dimension, and we can classify them as progressive or conservative based on their position on PC1.

```{r}
#| label: fig-PCA-WVS-statements
#| fig-cap: "1st Prinicple component for all statement based on World Values Survey data. "
gg <- sttmts |>
  arrange(desc(WVS_PC1)) |>
  mutate(Worldview_Statement_fct = fct_inorder(Worldview_Statement)) |>
  ggplot(aes(
    x = WVS_PC1,
    y = Worldview_Statement_fct,
    color = worldview,
    fill = WVS_PC1,
    tooltip = Worldview_Statement
  )) +
  geom_col_interactive() +
  labs(x = "", y = "", color = "Ideology LLM coded", fill = "WVS PC1") +
  scale_fill_continuous(palette = "RdYlBu", limits = c(-1, 1)) +
  scale_color_manual(values = c("blue", "red")) +
  theme_minimal(base_size = 8) +
  theme(axis.text.y = element_blank())
girafe(ggobj = gg, options = list(opts_toolbar(position = "top", fixed = TRUE)))
```


## Statement Popularity

@fig-statement-popularity shows how often each statement is part of a worldview at $t = 100$ over all simulation runs in all parameter settings. 

```{r}
#| label: fig-statement-popularity
#| fig-cap: "Popularity of statements at time step 100."
#| fig-width: 7.5
#| fig-asp: 0.9

## Popular Statements Sim at t=100
stats_pop <- memory |>
  filter(t == 100) |>
  left_join(runs_core, by = join_by(run_id)) |>
  # filter(openmindedness == 3, post == "F", M == 10) |>
  left_join(statements_meta, by = join_by(S_id)) |>
  count(run_id, S_id, Worldview_Statement, M, openmindedness, post) |>
  summarize(
    popularity = sum(n),
    .by = c(S_id, Worldview_Statement, M, openmindedness, post)
  )

stats_pop_total <- stats_pop |>
  summarize(popularity = sum(popularity), .by = c(S_id, Worldview_Statement)) |>
  mutate(label = glue("{S_id} {Worldview_Statement}")) |>
  arrange(desc(popularity)) |>
  left_join(sttmts |> select(S_id, WVS_PC1, worldview), by = join_by(S_id))
df <- stats_pop_total |>
  slice(1:100) |>
  mutate(
    group = if_else(row_number() <= 50, "Top 50", "51–100") |>
      factor(levels = c("Top 50", "51–100"))
  ) |>
  arrange(group, desc(popularity)) |>
  mutate(
    S_id_ordered = paste(S_id, group, sep = "___"),
    S_id_ordered = fct_inorder(S_id_ordered) |> fct_rev()
  )

pad <- 400 # data-unit space reserved for outside labels, same in both panels
p <- df |>
  ggplot(aes(
    y = S_id_ordered,
    x = popularity,
    label = popularity,
    tooltip = Worldview_Statement
  )) +
  geom_col_interactive(aes(
    color = worldview,
    fill = WVS_PC1,
  )) +
  geom_text(
    aes(
      hjust = ifelse(popularity > 1000, 1.1, -0.1)
    ),
    color = "black",
    size = 3
  ) +
  scale_y_discrete(labels = \(x) sub("___.*$", "", x)) +
  facet_wrap(~group, scales = "free", nrow = 1) +
  scale_x_continuous(
    breaks = c(),
    expand = expansion(mult = 0, add = c(0, pad))
  ) +
  scale_fill_continuous(palette = "RdYlBu", limits = c(-1, 1)) +
  scale_color_manual(values = c("blue", "red", "grey")) +
  labs(y = "", x = "", fill = "WVS PC1", color = "Ideology LLM coded") +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    legend.position = "inside",
    legend.position.inside = c(0.5, 0.5),
    legend.justification.inside = c(0.5, 0.5)
  )

ranges <- df |>
  group_by(group) |>
  summarise(xmax = max(popularity)) |>
  pull(xmax)
widths <- ranges + pad # matches the actual displayed range now, not an arbitrary offset

girafe(
  ggobj = p + force_panelsizes(cols = unit(widths, "null")),
  options = list(
    opts_toolbar(position = "top", fixed = TRUE)
  )
)
```




```{r}
#| label: fig-numstatements-per-run
#| tbl-cap: "Number of statements per run at t=100."

bubbles <- memory |>
  filter(t == 100) |>
  left_join(runs_core, by = join_by(run_id)) |>
  count(run_id_desc, M, openmindedness, post, S_id, seed)

bubbles_summary <- bubbles |>
  summarize(
    num_statements = n(),
    num_shared5_statements = sum(n >= 5),
    num_shared30_statements = sum(n >= 30),
    num_shared40_statements = sum(n > 40),
    num_shared50_statements = sum(n >= 50),
    num_shared70_statements = sum(n >= 70),
    .by = c(run_id_desc, M, openmindedness, post, seed)
  ) |>
  arrange(M, openmindedness, post)

# bubbles_summary |>
#   summarize(
#     avg_num_statements = mean(num_statements),
#     avg_num_shared5_statements = mean(num_shared5_statements),
#     avg_num_shared30_statements = mean(num_shared30_statements),
#     avg_num_shared40_statements = mean(num_shared40_statements),
#     avg_num_shared50_statements = mean(num_shared50_statements),
#     avg_num_shared70_statements = mean(num_shared70_statements),
#     .by = c(M, openmindedness, post)
#   ) |>
#   arrange(M, openmindedness, post) |>
#   gt() |>
#   fmt_number(decimal = 2) |>
#   opt_interactive(
#     use_sorting = TRUE,
#     use_highlight = TRUE,
#     use_compact_mode = TRUE,
#     use_text_wrapping = TRUE,
#     use_pagination = FALSE
#   ) |>
#   cols_width(
#     M ~ px(60),
#     openmindedness ~ px(120),
#     post ~ px(90)
#   )
```


## Number of surviving statements per run

The number of statements which exist in each run at $t=100$ is a simple indicator of how much a society has converged to a limited number of bubbles (here one or two) where agents hold very similar worldviews. An even stronger indicator is the number of statements which are shared by 40% of the agents or more. Both numbers are shown for all `r nrow(runs_core)` simulation runs in @fig-numstatements-per-run-plot.    

```{r}
#| label: fig-numstatements-per-run-plot
#| fig-cap: "Number of statements per simulation run at t=100. The top row of facets shows for all runs the total number of statements in the worldviews of all agents. The bottom row shows for the same runs the number of very popular statements which are shared by more than 40 agents (that means is 40% in all runs). Clicking on a point opens the network explorer for the corresponding run."
#| fig-height: 6
#| message: false
max_width <- 0.9 # total width available per category slot
step_max <- 0.18 # never space wider than this

gb_data <- bubbles_summary |>
  pivot_longer(
    c(num_statements, num_shared40_statements),
    names_to = "variable",
    values_to = "num"
  ) |>
  mutate(
    variable = fct_recode(
      variable,
      "#statements shared by >40 agents" = "num_shared40_statements",
      "#statements" = "num_statements"
    ) |>
      fct_rev(),
    onclick = paste0(
      'window.open("networkexplorer.html?net=',
      run_id_desc,
      '_t100")'
    ),
    xcat = interaction(openmindedness, post),
    x_num = as.numeric(xcat)
  ) |>
  mutate(
    n_tied = n(),
    step = pmin(step_max, max_width / pmax(n_tied - 1, 1)),
    x_pos = x_num + (row_number() - (n_tied + 1) / 2) * step,
    .by = c(M, variable, xcat, num)
  ) |>
  mutate(
    xcat = fct_recode(
      xcat,
      "Feed\nOp. 1" = "1.F",
      "Feed\nOp. 3" = "3.F",
      "Both\nOp. 1" = "1.FS",
      "Both\nOp. 3" = "3.FS",
      "Social\nOp. 1" = "1.S",
      "Social\nOp. 3" = "3.S"
    ),
    M = fct_recode(
      factor(M),
      "5 Statements per agent" = "5",
      "10 Statements per agent" = "10"
    )
  )

gb <- gb_data |>
  ggplot(
    aes(
      x = x_pos,
      y = num,
      fill = post,
      shape = variable,
      alpha = variable,
      tooltip = run_id_desc,
      onclick = onclick,
      data_id = run_id_desc
    )
  ) +
  geom_point_interactive(size = 1.6, stroke = 0.3) +
  scale_x_continuous(
    breaks = seq_along(levels(gb_data$xcat)),
    labels = levels(gb_data$xcat),
    minor_breaks = NULL
  ) +
  scale_y_continuous(minor_breaks = seq(0, 66, 2), breaks = seq(0, 66, 10)) +
  scale_shape_manual(values = c(21, 22)) +
  scale_fill_manual(values = c("red", "green", "blue")) +
  scale_alpha_manual(values = c(1, 0.5)) +
  #facet_wrap(~M) +
  facet_grid(variable ~ M, scales = "free_y") +
  labs(
    x = "Posting and Openmindedness",
    y = "",
    shape = "",
    fill = "Posting"
  ) +
  theme_minimal() +
  guides(alpha = "none", fill = "none", shape = "none") +
  theme(
    panel.border = element_rect(color = "black", fill = NA),
    legend.position = "bottom"
  )
girafe(
  ggobj = gb,
  options = list(
    opts_toolbar(position = "top", fixed = TRUE)
  )
)
```


# Conclusion

The LLM-based simulation match some findings from the original numerical model [@GeschkeEtAl2019TriplefilterBubbleUsing] qualitatively: 

- Bubbles of worldviews can emerge. These are groups of agents which share very similar statements while those statements are not oftenn shared by other agents. 
- Social posting spurs the formation of bubbles. Bubbles can also appear under feedposting alone but their their *bubbleness* is less pronounced. This demostrated by two trends in @fig-numstatements-per-run-plot: The number of surviving statements is lowest when only social posting is used while the number of statements shared by at least 40% of the agents is largest.  
- One bubble emerges slightly more often with higher openmindedness. This effect is less strong than in the numerical model. Likely because the LLMs do not consider our operationalization of openmindedness very strongly. 


*To do*


# Acknowledgements {.unnumbered}

We acknowledge the support of the European City Squared project, funded by the European Union's Horizon Europe research and innovation programme under grant agreement DOI: [10.3030/101178170](https://doi.org/10.3030/101178170). We also thank our colleagues in the project for their valuable feedback and discussions on this work.

We thank Alua Aldaniyaz and Asmae Nakib for their enlightening data analysis of World Value Survey data with the goal of quantifying worldviews of individuals in their student project on Visual Communication and Data Storytelling at Constructor University. Their work also built the basis of the selection of statements from the World Value Survey questionnaire for our model.

We thank Betül Yurtman for her work on assessing the empirical plausibility of emerging worldviews in the LLM-based agent-based model.  

# Appendix


## All Statements

```{r}
#| label: tbl-statements
#| tbl-cap: "All 100 statements. "

statements_meta |>
  select(WVS = Question_ID, S_id, Text = Worldview_Statement) |>
  mutate(
    WVS = WVS |>
      str_remove("Q") |>
      as.numeric()
  ) |>
  arrange(WVS) |>
  left_join(
    stats_pop_total |>
      select(S_id, Popularity = popularity),
    by = join_by(S_id)
  ) |>
  replace_na(list(Popularity = 0)) |>
  gt() |>
  opt_interactive(
    use_sorting = TRUE,
    use_filters = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 10,
    page_size_values = c(10, 25, 50, 100),
    pagination_type = c("numbers", "jump", "simple")
  ) |>
  cols_width(
    WVS ~ px(60),
    S_id ~ px(90),
    Text ~ px(500),
    Popularity ~ px(100)
  )
```

## Eye-balled network types

Before turning to a mathematical network metrics to quantify different types we made a manual coding using eye-bulling on three different aspects: *number of bubbles* quantifies if the network is better quantified to be 1, 2, or 3 bubbles. A *bubble* is similar to what is also called a module or a cluster. It is supposed to be a group of nodes which are more connected among each other than to the other nodes. In our manual coding a we consider a diverse network without a clear bubble structure to be one bubble. As this classification can not distinguish between tightly connected bubbles and wide stread structures, the next aspect is bubbleness. *Bubbleness* has a value of 0, 0.25, 0.5, 0.75, or 1 where 1 means that all bubbles have not many statements and are therefore tightly connected and 0 means that the number of statements is still very large and bubble structure if not visible. Finally, *parity* takes also the values 0, 0.25, 0.5, 0.75, or 1. A network with one bubble always receives the value 0. For networks with two or three bubbles 1 specifies that the bubbles are almost equal in size with respect to the number of agents, while 0 stands for a network where one of the bubbles includes almost all nodes.

So, the clearest indicator of a consensusal society is with one bubble with maximal bubbleness (1, 1, 0). A maximally bipolarized polarized soceity has two bubbles with maximal bubbleness and parity. is most clearly quantified with (2, 1, 1).  

```{r}
#| label: tbl-network-types
#| tbl-cap: "All simulation runs"

statements_count <- memory |>
  filter(t == 100) |>
  count(run_id, S_id) |>
  summarize(
    num_statements = n(),
    num_shared5_statements = sum(n >= 5),
    num_shared30_statements = sum(n >= 30),
    num_shared50_statements = sum(n >= 50),
    num_shared70_statements = sum(n >= 70),
    .by = run_id
  )

runs_analysis <- runs_core |>
  left_join(
    runs_annotations |> select(run_id, num_bub, bubbleness, parity),
    by = join_by(run_id)
  ) |>
  left_join(statements_count, by = join_by(run_id)) |>
  select(
    M,
    Openminded = openmindedness,
    Posting = post,
    Seed = seed,
    num_statements,
    num_shared5_statements,
    num_shared30_statements,
    num_shared50_statements,
    num_shared70_statements,
    Number_Bubbles = num_bub,
    Bubbleness = bubbleness,
    Parity = parity,
    Filename = run_id_desc
  )
runs_analysis |>
  gt() |>
  opt_interactive(
    use_sorting = TRUE,
    use_filters = TRUE,
    use_highlight = TRUE,
    use_compact_mode = TRUE,
    use_text_wrapping = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 10,
    page_size_values = c(10, 25, 72),
    pagination_type = c("numbers", "jump", "simple")
  ) |>
  cols_width(
    M ~ px(60),
    Openminded ~ px(120),
    Posting ~ px(90),
    Seed ~ px(90),
    Number_Bubbles ~ px(140),
    Bubbleness ~ px(100),
    Parity ~ px(90),
    Filename ~ px(500)
  )
```
